Quantum Mechanical Many-body Problems with Machine Learning Algorithms

### 2019-03-19 doc/MachineLearning/article2.tex
update

Traditional many-body methods like full configuration interaction
theory (FCI), coupled-cluster (CC) theory, many-body perturbation
theory (MBPT), in-medium similarity renormalization group (IMSRG),
various Monte Carlo methods and many other many-body approaches, have
been rather successful in describing properties of interacting
many-body systems. These methods have been widely used in condensed
matter physics, nuclear physics, quantum chemistry and materials
science, just to mention a few of the areas of applicability.

However, essentially all of these methods face what is
normally called the curse of dimensionality. For wave function based
method like FCI or CC theories where the original continuous problem
is discretized in terms selected basis functions and/or many-body
excitations, the dimensionality of the systems under study grows almost
exponentially when larger basis sets and/or number of particles are
included. As an example, nuclear physics 
systems close to the limits of stability pose in
particular a tough problem to many-body practitioners in terms of a dramaticn
increase of the number of relevant degrees of freedom. To describe say
weakly bound nuclei within the framework of a wave function based
approach, requires often a single-particle basis which includes bound,
weakly bound and unbound states, increasing thereby considerably the
number of many-body basis states. This renders often a standard
FCI calculation infeasible. Within the above mentioned
many-body methods there are however several interesting theoretical
approachs which attempt at circumventing the dimensionality
curse. Smarter basis sets is one of these approaches, as well as the
resummation of specific correlations and the recently proposed
stochastic sampling of many-body states in both FCI and CC
calculations \cite{alavi2009,thom2017}.

Recently, several authors have pointed to  possibilities within the
broad fields of Machine Learning and Quantum Computing as approaches
that hold great promise in tackling the ever increasing
dimensionalities. There are several groups worldwide which now focus
on Machine Learning and/or Quantum Computing applied to many-body
problems.

This work is inspired by the idea of representing the wave function with
a restricted Boltzmann machine (RBM), presented recently by \href{{http://science.sciencemag.org/content/355/6325/602}}{G. Carleo and M. Troyer, Science \textbf{355}, Issue 6325, pp. 602-606 (2017)}. They
named such a wave function/network a \emph{neural network quantum state} (NQS). In their article they apply it to the quantum mechanical
spin lattice systems of the Ising model and Heisenberg model, with
encouraging results.

Here we try to implement these algorithms to many-body systems that mimick a nuclear physics like problem, namely electrons confined to move in harmonic oscillator like potentials.

Machine learning (ML) is an extremely rich field, in spite of its
young age. The increases we have seen during the last three decades in
computational capabilities have been followed by developments of
methods and techniques for analyzing and handling large date sets,
relying heavily on statistics, computer science and mathematics.  The
field is rather new and developing rapidly.

@@

\section{What is Machine Learning}
Machine learning is the science of giving computers the ability to
learn without being explicitly programmed.  The idea is that there
exist generic algorithms which can be used to find patterns in a broad
class of data sets without having to write code specifically for each
problem. The algorithm will build its own logic based on the data.

>>>>

Traditional many-body methods like full configuration interaction
theory (FCI) \cite{fcirefs}, coupled-cluster (CC) theory
\cite{ccrefs}, many-body perturbation theory (MBPT) \cite{mbptrefs},
in-medium similarity renormalization group (IMSRG) \cite{imsrgrefs},
various Monte Carlo methods \cite{mcrefs} and many other many-body
approaches, have been rather successful in describing properties of
interacting many-body systems. These methods have been widely used in
condensed matter physics, nuclear physics, quantum chemistry and
materials science, just to mention a few of the areas of
applicability.

However, essentially all of these methods face what is
normally called the curse of dimensionality. For wave function based
method like FCI or CC theories where the original continuous problem
is discretized in terms selected basis functions and/or many-body
excitations, the dimensionality of the systems under study grows almost
exponentially when larger basis sets and/or number of particles are
included. As an example, nuclear physics 
systems close to the limits of stability \cite{nplims} pose in
particular a tough problem to many-body practitioners in terms of a dramaticn
increase of the number of relevant degrees of freedom. To describe say
weakly bound nuclei within the framework of a wave function based
approach, requires often a single-particle basis which includes bound,
weakly bound and unbound states, increasing thereby considerably the
number of many-body basis states. This renders often a standard
FCI calculation infeasible. Within the above mentioned
many-body methods there are however several interesting theoretical
approachs which attempt at circumventing the dimensionality
curse. Smarter basis sets is one of these approaches, as well as the
resummation of specific correlations and the recently proposed
stochastic sampling of many-body states in both FCI and CC
calculations \cite{alavi2009,thom2017}.

Recently, several authors have pointed to  possibilities within the
broad fields of Machine Learning and Quantum Computing as approaches
that hold great promise in tackling the ever increasing
dimensionalities. There are several groups worldwide which now focus
on Machine Learning and/or Quantum Computing applied to many-body
problems \cite{CarleoTroyer2017}.
Machine learning (ML) is an extremely rich field, in spite of its
young age. The increases we have seen during the last three decades in
computational capabilities have been followed by developments of
methods and techniques for analyzing and handling large date sets,
relying heavily on statistics, computer science and mathematics.  The
field is rather new and developing rapidly.

Within this setting, the aim of this work is to explore similarities
and differences between widely used many-body methods like the
Variational Monte Carlo or the diffusion Monte Carlo approaches and
unsupervised Machine Learning methods using so-called Boltzmann
machines. We will also point to similarities between Monte Carlo
approaches with shadow wave functions and Machine Learning algorithms
that employ Boltzmann machines. Since the field of Machine Learning is
evolving rather rapidly and physicists may not be fully familiar with
the plethora of Machine Learning algorithms, we will try in this
work to keep a more pedestrian approach. In particular, we will try to outline 
some of the similarities between Monte Carlo methods for solving quantum mechanical many-body problems and Machine Learning algorithms.

We start thus with a brief review of central aspects of Monte Carlo
methods in the next section. Our model system is that of electrons
confined to move in two or three dimensions, so-called quantum
dots. We will in particular focus on idealized quantum dots with
electrons moving in a harmonic oscillator potential. There are several
reasons for focusing on quantum dots. For two interacting electrons
there are analytical eigenpairs, allowing us thereby to benchmark
various many-body methods \cite{taut1993,taut1994}. For more than two
electrons, there are extensive reference calculations based on a
variety of many-body methods. These calculations, in addition to new
ones presented here, will allow us to assess the relevance of our
results obtained with Machine Learning algorithms.

Thereafter, in Section \ref{sec:MLmethods}, we present and discuss the implementation of Machine Learning algorithms. In that section we present also a discussion of possible similarities between Boltzmann machines and so-called Shadow wave functions.

\section{What is Machine Learning}

@@

Machine learning is the science of giving computers the ability to
learn without being explicitly programmed.  The idea is that there
exist generic algorithms which can be used to find patterns in a broad
class of data sets without having to write code specifically for each
problem. The algorithm will build its own logic based on the data.
